---
title: 'Are elevated lactate levels associated with ICU mortality?'
author: "..."
date: "January, 2025"
output:
  pdf_document: 
    toc: true
    toc_depth: 4
    number_sections: true
    fig_width: 7
    fig_height: 5
    fig_caption: true
    keep_tex: false
    latex_engine: xelatex
    highlight: tango
    citation_package: natbib
header-includes:
- \usepackage{xcolor}
- \usepackage{framed}
- \usepackage{listings}
- \usepackage{fancyhdr}
- \usepackage{geometry}
- \usepackage{caption}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{graphicx}
- \geometry{top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm}
- \pagestyle{fancy}
- \fancyfoot[C]{\thepage}
- \renewcommand{\headrulewidth}{0.4pt}
- \renewcommand{\footrulewidth}{0.4pt}
editor_options: 
  markdown: 
    wrap: 72
---

\newpage

Authors' information

\begin{center}
\begin{tabular}{|l|l|}
\hline
\textbf{Nombre} & Johanna Ursula Albers \\
\hline
\textbf{Email} & johanna.albers@estudiants.urv.cat\\
\hline
\textbf{GitHub} & https://github.com/ \\
\hline
\textbf{LinkedIn} & https://www.linkedin.com/in/ \\
\hline
\textbf{Fecha} & \today \\
\hline
\end{tabular}
\end{center}
\begin{center}
\begin{tabular}{|l|l|}
\hline
\textbf{Nombre} & Marco Russo \\
\hline
\textbf{Email} & mrussorb@uoc.edu \\
\hline
\textbf{GitHub} & https://github.com/marcusRB/ \\
\hline
\textbf{LinkedIn} & https://www.linkedin.com/in/marcusrb/ \\
\hline
\textbf{Fecha} & \today \\
\hline
\end{tabular}
\end{center}
\begin{center}
\begin{tabular}{|l|l|}
\hline
\textbf{Nombre} & Noa-Sibila Janer Oliver \\
\hline
\textbf{Email} & noa-sibila.janer@estudiants.urv.cat \\
\hline
\textbf{GitHub} & https://github.com/ \\
\hline
\textbf{LinkedIn} & https://www.linkedin.com/in/ \\
\hline
\textbf{Fecha} & \today \\
\hline
\end{tabular}
\end{center}
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

# Project overview

**Research Question**: Are elevated lactate levels associated with ICU
mortality? Study Design: Retrospective cohort study using MIMIC-III
database Primary Analysis: Association between first lactate measurement
post-ICU admission and in-hospital mortality.

## Data Description

MIMIC-III is a relational database consisting of 26 tables. Tables are
linked by identifiers which usually have the suffix ‘ID’. For example,
SUBJECT_ID refers to a unique patient, HADM_ID refers to a unique
admission to the hospital, and ICUSTAY_ID refers to a unique admission
to an intensive care unit.

Charted events such as notes, laboratory tests, and fluid balance are
stored in a series of ‘events’ tables. For example the OUTPUTEVENTS
table contains all measurements related to output for a given patient,
while the LABEVENTS table contains laboratory test results for a
patient.

Tables prefixed with ‘D\_’ are dictionary tables and provide definitions
for identifiers. For example, every row of CHARTEVENTS is associated
with a single ITEMID which represents the concept measured, but it does
not contain the actual name of the measurement. By joining CHARTEVENTS
and D_ITEMS on ITEMID, it is possible to identify the concept
represented by a given ITEMID.

Developing the MIMIC data model involved balancing simplicity of
interpretation against closeness to ground truth. As such, the model is
a reflection of underlying data sources, modified over iterations of the
MIMIC database in response to user feedback. Care has been taken to
avoid making assumptions about the underlying data when carrying out
transformations, so MIMIC-III closely represents the raw hospital data.

Broadly speaking, five tables are used to define and track patient
stays: ADMISSIONS; PATIENTS; ICUSTAYS; SERVICES; and TRANSFERS. Another
five tables are dictionaries for cross-referencing codes against their
respective definitions: D_CPT; D_ICD_DIAGNOSES; D_ICD_PROCEDURES;
D_ITEMS; and D_LABITEMS. The remaining tables contain data associated
with patient care, such as physiological measurements, caregiver
observations, and billing information.

In some cases it would be possible to merge tables—for example, the
D_ICD_PROCEDURES and CPTEVENTS tables both contain detail relating to
procedures and could be combined—but our approach is to keep the tables
independent for clarity, since the data sources are significantly
different. Rather than combining the tables within MIMIC data model, we
suggest researchers develop database views and transforms as
appropriate.

## MIMIC-III v1.4

The current version of the database is v1.4. When referencing this
version, we recommend using the full title: MIMIC-III v1.4[^1].

[^1]: Johnson, A., Pollard, T., & Mark, R. (2016). MIMIC-III Clinical
    Database (version 1.4). PhysioNet. RRID:SCR_007345.
    <https://doi.org/10.13026/C2XW26>

MIMIC-III v1.4 was released on 2 September 2016. It was a major release
enhancing data quality and providing a large amount of additional data
for Metavision patients.

***

\newpage

# Data Extraction & Preparation

## Database setup, cohort extraction

```{r database setup, include=FALSE}
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("tidyr")) install.packages("tidyr")
if (!require("dplyr")) install.packages("dplyr")
if (!require("tibble")) install.packages("tibble"())
if (!require("lubridate")) install.packages("lubridate")
if (!require("readr")) install.packages("readr")
if (!require("stringr")) install.packages("stringr")
if (!require("ggplot2")) install.packages("ggplot2")
library(tidyverse)
library(dplyr)
library(tidyr)
library(tibble)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)

if (!require("data.table")) install.packages("data.table")
if (!require("odbc")) install.packages("odbc")
if (!require("RMariaDB")) install.packages("RMariaDB")
if (!require("broom")) install.packages("broom")
if (!require("DescTools")) install.packages("DescTools")
if (!require("boot")) install.packages("boot")
if (!require("effsize")) install.packages("effsize")
library(data.table)
library(odbc)
library(RMariaDB)
library(tidyverse)
library(broom)
library(DescTools)
library(boot)
library(effsize)

if (!require("cluster")) install.packages("cluster")
if (!require("factoextra")) install.packages("factoextra")
if (!require("NbClust")) install.packages("NbClust")
if (!require("dbscan")) install.packages("dbscan")
if (!require("mclust")) install.packages("mclust")
library(cluster)
library(factoextra)
library(NbClust)
library(dbscan)
library(mclust)

if (!require("caret")) install.packages("caret")
if (!require("C50")) install.packages("C50")
if (!require("randomForest")) install.packages("randomForest")
if (!require("lightgbm")) install.packages("lightgbm")
if (!require("pROC")) install.packages("pROC")
if (!require("DALEX")) install.packages("DALEX")
if (!require("DALEXtra")) install.packages("DALEXtra")
if (!require("shapviz")) install.packages("shapviz")
if (!require("kernelshap")) install.packages("kernelshap")

library(caret)
library(C50)
library(randomForest)
library(lightgbm)
library(pROC)
library(DALEX)
library(DALEXtra)
library(shapviz)
library(kernelshap)
```


```{r}
# Creating the main directories
if (!dir.exists("figures")) {
dir.create("figures", recursive = TRUE)
}

if (!dir.exists("results")) {
dir.create("results", recursive = TRUE)
}

```



```{r include=FALSE}
# Get the credentials file
# It looks like:
################
#username=your.user.name
#password=your.password.assigned
################
creds_file = "creds.txt"
```


```{r include=FALSE}
# Set the username and password
creds <- readLines(creds_file)
cred_list <- setNames(
  sub(".*=", "", creds),
  sub("=.*", "", creds)
)
```

```{r include=FALSE}
con <- dbConnect(
  drv = RMariaDB::MariaDB(),
  username = cred_list[["username"]],
  password = cred_list[["password"]],
  host = "ehr3.deim.urv.cat",
  dbname = "mimiciiiv14",
  port = 3306
)
```

```{r, results='hide'}
#| echo: false
dbListTables(con)
```

We constructed a cohort at the level of individual ICU stays. Each row
in the dataset corresponds to a single ICU stay. The most important
measurement, the first lactate measurement after the ICU admission is
included.

Lactate measurements were assigned to ICU stays based on their
timestamp: a measurement was linked to an ICU stay only if it occurred
between the ICU admission (intime) and 24h after.

ICU mortality was derived from hospital admission outcomes. For patients
with multiple ICU stays within the same hospital admission, ICU
mortality was assigned exclusively to the final ICU stay if the patient
died during that admission; all preceding ICU stays were labeled as
non-fatal. This approach allows ICU-level mortality attribution in the
absence of a direct ICU death indicator. The final cohort comprised
129,966 lactate measurements, representing multiple ICU stays and
hospital admissions across patients.

Further confounding variables are also included: age, gender, creatine
kinase, bilirubin and mean arterial blood pressure (map). The inlcuded
measurements are the first measurements in the first 24 h after the ICU
admission.

```{r include=FALSE}
library(DBI)
dbExecute(con, "SET SQL_BIG_SELECTS=1;")
```

```{r include=FALSE}
cohort_query <- "
        WITH icu_ordered AS (
        SELECT
            i.subject_id,
            i.hadm_id,
            i.icustay_id,
            i.intime,
            i.outtime,
            a.hospital_expire_flag AS hospital_dead,
            ROW_NUMBER() OVER (
                PARTITION BY i.subject_id, i.hadm_id
                ORDER BY i.outtime DESC
            ) AS rn_icu
        FROM ICUSTAYS i
        JOIN ADMISSIONS a
            ON i.hadm_id = a.hadm_id
    ),
    lactate_first AS (
        SELECT
            io.subject_id,
            io.hadm_id,
            io.icustay_id,
            l.charttime as lactate_time,
            l.value as lactate_value,
            l.valuenum as lactate_value_num,
            l.valueuom as lactate_units,
            l.flag as lactate_flag,
            ROW_NUMBER() OVER (
                PARTITION BY io.icustay_id
                ORDER BY l.charttime ASC
            ) AS rn
        FROM LABEVENTS l
        INNER JOIN icu_ordered io USING(subject_id, hadm_id)
        WHERE l.itemid = 50813
          AND l.value IS NOT NULL
          AND l.charttime BETWEEN io.intime AND (io.intime + INTERVAL '24' HOUR)
    ),
    ck_first AS (
        SELECT
            io.subject_id,
            io.hadm_id,
            io.icustay_id,
            l.charttime as ck_time,
            l.value as ck_value,
            l.valuenum as ck_value_num,
            l.valueuom as ck_units,
            l.flag as ck_flag,
            ROW_NUMBER() OVER (
                PARTITION BY io.icustay_id
                ORDER BY l.charttime ASC
            ) AS rn
        FROM LABEVENTS l
        INNER JOIN icu_ordered io USING(subject_id, hadm_id)
        WHERE l.ITEMID = 50910
          AND l.charttime BETWEEN io.intime AND (io.intime + INTERVAL '24' HOUR)
    ),
    bilirubin_first AS (
        SELECT
            io.subject_id,
            io.hadm_id,
            io.icustay_id,
            l.charttime as bilirubin_time,
            l.value as bilirubin_value,
            l.valuenum as bilirubin_value_num,
            l.valueuom as bilirubin_units,
            l.flag as bilirubin_flag,
            ROW_NUMBER() OVER (
                PARTITION BY io.icustay_id
                ORDER BY l.charttime ASC
            ) AS rn
        FROM LABEVENTS l
        INNER JOIN icu_ordered io USING(subject_id, hadm_id)
        WHERE l.ITEMID = 50885
          AND l.charttime BETWEEN io.intime AND (io.intime + INTERVAL '24' HOUR)
    ),
    mabp_first AS (
        SELECT
            io.subject_id,
            io.hadm_id,
            io.icustay_id,
            c.CHARTTIME as mabp_time,
            c.value as mabp_value,
            c.valuenum as mabp_value_num,
            c.valueuom as mabp_units,
            ROW_NUMBER() OVER (
                PARTITION BY io.icustay_id
                ORDER BY c.charttime ASC
            ) AS rn
        FROM CHARTEVENTS c
        INNER JOIN icu_ordered io USING(subject_id, hadm_id, icustay_id)
        WHERE c.itemid = 220052
          AND c.valuenum IS NOT NULL
          AND c.charttime BETWEEN io.intime AND (io.intime + INTERVAL '24' HOUR)
    ),
    icu_with_lactate AS (
        SELECT DISTINCT subject_id, hadm_id, icustay_id
        FROM lactate_first
        WHERE rn = 1
    )
    SELECT
        ROW_NUMBER() OVER (ORDER BY iu.subject_id, iu.hadm_id, iu.icustay_id) AS row_id,
        iu.subject_id,
        p.dob,
        YEAR(iu.intime) - YEAR(p.DOB) AS age_icu_entry,
        YEAR(iu.outtime) - YEAR(p.DOB) AS age_icu_exit,
        p.gender,
        iu.hadm_id,
        iu.icustay_id,
        iu.intime,
        iu.outtime,
        iu.hospital_dead,
        CASE
            WHEN iu.hospital_dead = 1 AND iu.rn_icu = 1 THEN 1
            ELSE 0
        END AS icu_dead,
        lf.lactate_time,
        lf.lactate_value,
        lf.lactate_value_num,
        lf.lactate_units,
        lf.lactate_flag,
        ck.ck_time,
        ck.ck_value,
        ck.ck_value_num,
        ck.ck_units,
        ck.ck_flag,
        bil.bilirubin_time,
        bil.bilirubin_value,
        bil.bilirubin_value_num,
        bil.bilirubin_units,
        bil.bilirubin_flag,
        mb.mabp_time,
        mb.mabp_value,
        mb.mabp_value_num,
        mb.mabp_units
    FROM icu_ordered iu
    INNER JOIN icu_with_lactate iwl ON iwl.subject_id = iu.subject_id 
        AND iwl.hadm_id = iu.hadm_id 
        AND iwl.icustay_id = iu.icustay_id
    LEFT JOIN lactate_first lf ON lf.subject_id = iu.subject_id 
        AND lf.hadm_id = iu.hadm_id 
        AND lf.icustay_id = iu.icustay_id
        AND lf.rn = 1
    LEFT JOIN ck_first ck ON ck.subject_id = iu.subject_id 
        AND ck.hadm_id = iu.hadm_id 
        AND ck.icustay_id = iu.icustay_id
        AND ck.rn = 1
    LEFT JOIN bilirubin_first bil ON bil.subject_id = iu.subject_id 
        AND bil.hadm_id = iu.hadm_id 
        AND bil.icustay_id = iu.icustay_id
        AND bil.rn = 1
    LEFT JOIN mabp_first mb ON mb.subject_id = iu.subject_id 
        AND mb.hadm_id = iu.hadm_id 
        AND mb.icustay_id = iu.icustay_id
        AND mb.rn = 1
    LEFT JOIN PATIENTS p ON p.subject_id = iu.subject_id
    ORDER BY iu.subject_id, iu.hadm_id, iu.icustay_id;
"
```


```{r}
initial_df <- dbGetQuery(con, cohort_query) 
initial_df %>% glimpse()
```


```{r include=FALSE}
# Disconnect from the database
dbDisconnect(con)
```


***

## Data cleaning

First we checked the data for duplicated. No duplicates were found.

```{r}
# Are there duplicates?
sum(duplicated(initial_df))
```

Checking the initial dataframe, we saw some null values to analyse if to impute
or to remove in the next steps.

```{r}
as.data.frame(initial_df) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  select(where(~ . > 0))
```

We only kept observations, which are greater then 0 mmol/L, made sure
that all lactate values are numeric and checked for suspicious values.

```{r}
clean_values_df <- initial_df %>%
  mutate(
    lactate_value_num = case_when(
      # For samples containing ">" or "GREATER THAN 30" assign value 31
      
      str_detect(lactate_value, fixed(">")) | 
      str_detect(toupper(lactate_value), "GREATER THAN") ~ 31,
      
      TRUE ~ as.numeric(lactate_value) # Convert to numeric value, 
      # samples with numbers and letters will be converted to NA
    )
  ) %>%
  filter(!is.na(lactate_value_num)) # Delete whole row if lactate value is null


# Check for suspicious values
clean_values_df %>%
  filter(lactate_value_num <= 0 | lactate_value_num > 31)

```

We verified the logical consistency of mortality indicators by
confirming that ICU death did not occur in the absence of in-hospital
death.

```{r}
clean_values_df %>%
  filter(icu_dead == 1 & hospital_dead == 0) %>%
  nrow()
```

We assessed mortality consistency at the patient level by confirming
that each subject had at most one recorded ICU death across all ICU
stays.

```{r}
death_counts <- clean_values_df %>%
  group_by(subject_id) %>%
  summarise(
    hosp_dead = n_distinct(hadm_id[hospital_dead == 1], na.rm = TRUE),
    icu_dead  = n_distinct(icustay_id[icu_dead == 1], na.rm = TRUE)
  ) %>%
  filter(icu_dead > 1)
death_counts
```

A data inconsistency was identified involving a single subject with two
incorrectly assigned mortality flags; these records were manually
rectified to ensure data integrity.

```{r}
clean_values_df <- clean_values_df %>%
  mutate(
    # Reassign hospital_dead from first admission
    hospital_dead = if_else(subject_id == 17796 & hadm_id == 119823, 0, hospital_dead),
    
    # Reassign icu_dead from first ICU stay
    icu_dead = if_else(subject_id == 17796 & hadm_id == 119823, 0, icu_dead)
  )
```

For our analysis, we will only need the following columns: `icustay_id`,
`age_icu_exit`, `gender`, `icu_dead`, `lactate_value_num`, `ck_value_num`,
`bilirubin_value_num`, `mabp_value_num.` Regarding the age, we chose the
attribute exiting the ICU, because it is the age of death or survival. These
columns will be defined in *analysis_df*.

Laboratory abnormality flags were excluded from analysis, as continuous
laboratory measurements provide greater informational content and allow
more precise modeling than qualitative classifications.

```{r}
clean_values_df <- clean_values_df %>%
  select(
    icustay_id, age_icu_exit, gender, icu_dead, lactate_value_num, ck_value_num, bilirubin_value_num, mabp_value_num
  )

head(clean_values_df)
```

```{r}
summary(clean_values_df)
```

After the first step of the data cleaning, reducing the variables in 8 from 31.

Creatine kinase (CK) and bilirubin values were inspected for
plausibility. Although both variables showed highly skewed distributions
and substantial missingness, extreme values were considered clinically
plausible (e.g., severe rhabdomyolysis for CK and advanced liver failure
for bilirubin). Therefore, no values were removed during data cleaning.

Due to de-identification procedures in MIMIC-III, patients aged 89 years
or older have artificially shifted dates of birth. Therefore me have a
maximum of 311 in the column age_icu_exit. Age values exceeding 89 years
were therefore capped at 90 years for analysis.

```{r}
clean_values_df <- clean_values_df %>%
  mutate(
    age_icu_exit = if_else(age_icu_exit > 89, 90, age_icu_exit)
  )
```

Physiologically implausible mean arterial blood pressure values (\<30 or
\>200 mmHg) were set to missing. 

This will be our final step before the EDA, so we will simplify the name
of the data frame to `eda_df`.

```{r}
eda_df <- clean_values_df %>%
  mutate(
    mabp_value_num = if_else(
      mabp_value_num < 30 | mabp_value_num > 200,
      NA_real_,
      mabp_value_num
    )
  )
```

***

\newpage

## Exploratory analysis

We start showing the shape and short summary of the dataframe.

```{r}
dim(eda_df)
```

```{r}
summary(eda_df)
```

The data frame *comprises 22,302 ICU stays*, including lactate
observations, gender and age, providing a robust longitudinal
perspective on patient biomarkers.

We describe which patients are in our data set.

```{r}
table(eda_df$gender)
table(clean_values_df$icu_dead)
prop.table(table(clean_values_df$icu_dead))*100
```

We analyse the distribution of the lactate value.

```{r}
# Show the distribution of lactate values
figure_1 <- ggplot(eda_df, aes(lactate_value_num)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") + 
  labs(title = "Distribution of lactate values") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
figure_1
```


```{r include=FALSE}
ggsave(
  filename = "../figures/figure01_distribution_lactate_value.png",
  plot = figure_1,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```


We add a column with the log-transformation of the lactate measurement
because the values are skewed.

```{r}
eda_df <- eda_df %>%
  mutate(
    log_lactate = log(lactate_value_num)
  )
```


```{r}
# Show the distribution of lactate values (log scale)
figure_2 <- ggplot(eda_df, aes(log_lactate)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") + 
  labs(title = "Distribution of lactate values (log scale)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
figure_2
```

```{r}
ggsave(
  filename = "../figures/figure02_distribution_lactate_value_log.png",
  plot = figure_2,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```



We analyse the distribution of confounding variables.

```{r}
# Show the distribution of bilirubin values
figure_3 <- ggplot(eda_df, aes(bilirubin_value_num)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") + 
  labs(title = "Distribution of creatine bilirubin values") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

figure_3
```

```{r}
ggsave(
  filename = "../figures/figure03_distribution_bilirubin_value.png",
  plot = figure_3,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```



```{r}
# Show the distribution of ck values
figure_4 <- ggplot(eda_df, aes(ck_value_num)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") + 
  labs(title = "Distribution of creatine kinase values") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
figure_4
```

```{r}
ggsave(
  filename = "../figures/figure04_distribution_ck_value.png",
  plot = figure_4,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```



Although creatine kinase and bilirubin measurements exhibited extreme
high values, observed concentrations were within clinically plausible
ranges and were therefore retained. The values exhibited extreme right
skew and were therefore log-transformed. 0,01 was added to
bilirubin_value_num in case bilirubin_value_num = 0.

```{r}
eda_df <- eda_df %>%
  mutate(
    log_ck = log(ck_value_num),
    log_bilirubin = log(bilirubin_value_num + 0.01)
  )
```

```{r}
# Show the distribution of ck values (log scale)
figure_5 <- ggplot(eda_df, aes(log_ck)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") + 
  labs(title = "Distribution of ck values (log scale)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
figure_5
```

```{r}
ggsave(
  filename = "../figures/figure05_distribution_ck_value_log.png",
  plot = figure_5,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```



```{r}
# Show the distribution of bilirubin values (log scale)
figure_6 <- ggplot(eda_df, aes(log_bilirubin)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") + 
  labs(title = "Distribution of bilirubin values (log scale)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
figure_6
```


```{r}
ggsave(
  filename = "../figures/figure06_distribution_bilirubin_log.png",
  plot = figure_6,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```




```{r}
# Show the distribution of mabp values
figure_7 <- ggplot(eda_df, aes(mabp_value_num)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") + 
  labs(title = "Distribution of mabp values") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
figure_7
```



```{r}
ggsave(
  filename = "../figures/figure07_distribution_mabp_values.png",
  plot = figure_7,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```




We examined the association between log-transformed lactate levels and
ICU mortality, which constituted the primary relationship of interest.

```{r}
figure_08 <- ggplot(eda_df, aes(x = factor(icu_dead), y = log_lactate)) +
  geom_boxplot() +
  labs(
    x = "ICU Death",
    y = "Log(Lactate)"
  )

figure_08
```

```{r}
ggsave(
  filename = "../figures/figure08_boxplot_icu_death.png",
  plot = figure_08,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```

***

### Missing values and imputations

Checking the missing values


```{r}
library(naniar)
library(ggplot2)
# Visualize missing values by variable

missing_df <- eda_df %>%
  select(-icustay_id)

figure09 <- gg_miss_var(missing_df, show_pct = TRUE) +
labs(title = "Missing Values by Variable in Dataset Subset",
x = "Variables",
y = "Proportion of Missing Values")
figure09
```

```{r}
ggsave(
  filename = "../figures/figure09_missing_values.png",
  plot = figure09,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```


```{r}
# Check missing values
missing_summary <- sapply(eda_df, function(x) sum(is.na(x)))
missing_df <- data.frame(
    Column = names(missing_summary),
    Missing_Count = missing_summary,
    Missing_Percent = round(missing_summary/nrow(eda_df)*100, 2)
  ) %>%
  arrange(desc(Missing_Count))

print(missing_df)
```

Several covariates exhibited substantial missingness, most notably mean 
arterial blood pressure (73.7%), creatine kinase (58.8%), and bilirubin (56.6%).
As a result, the inclusion of these variables as confounders in multivariable
models was limited. However, the primary exposure (lactate) and 
outcome (ICU mortality), as well as key demographic variables, were complete.

Missingness was documented and explored during exploratory analysis, 
and no imputation was performed.

*** 

## Correlations / collinearity

```{r}
cor(
  eda_df %>%
    select(log_lactate,
           mabp_value_num,
           log_bilirubin,
           log_ck),
  #Only use rows where all selected variables are observed (no NA values)
  use = "complete.obs"
)
```

Correlations among continuous predictors were weak, indicating no evidence of problematic collinearity.

***

## EDA Summary

Based on exploratory data analysis, lactate was retained as the primary exposure variable and modeled on the log scale due to right-skewed distributions. Creatine kinase and bilirubin were likewise log-transformed prior to analysis. Mean arterial blood pressure was retained on the original scale after removal of physiologically implausible values. Laboratory abnormality flags were excluded, as quantitative measurements provided greater informational content. Correlation analysis indicated no problematic collinearity among continuous predictors. Variables with substantial missingness were considered as secondary confounders, and no imputation was performed.

***

\newpage

# Statistical Analysis

We define the data frame for this part as *analysis_df*.

```{r}
analysis_df <- eda_df %>%
  select(
    icustay_id, age_icu_exit, gender, icu_dead,
    log_lactate, log_ck, log_bilirubin, mabp_value_num,
    bilirubin_value_num, ck_value_num, lactate_value_num
  )

head(analysis_df)
```

## Fit the main (core) model

```{r}
model_main <- glm(
  icu_dead ~ log_lactate + age_icu_exit + gender,
  data = analysis_df,
  family = binomial
)
summary(model_main)
```
In multivariable logistic regression adjusted for age and gender, higher log-transformed lactate levels were strongly associated with increased ICU mortality (OR ≈ 2.8 per unit increase in log-lactate, p < 0.001). Age was independently associated with mortality, while no significant association with gender was observed.

## Fit the model with all variables

This tests robustness to confounding.

```{r}
model_adjusted <- glm(
  icu_dead ~ log_lactate + age_icu_exit + gender +
             mabp_value_num + log_bilirubin + log_ck,
  data = analysis_df,
  family = binomial
)
summary(model_adjusted)
```
After adjustment for age, gender, mean arterial blood pressure, bilirubin, and creatine kinase, higher log-transformed lactate levels remained strongly associated with increased ICU mortality (OR ≈ 2.6 per unit increase in log-lactate).

MAP was not independently associated with ICU mortality after adjustment.

Higher bilirubin levels were independently associated with increased ICU mortality, consistent with liver dysfunction severity.

No independent association between creatine kinase and ICU mortality was observed after adjustment.

In a multivariable logistic regression adjusted for age and gender, higher log-transformed lactate levels were strongly associated with increased ICU mortality (OR ≈ 2.8). This association remained robust after additional adjustment for mean arterial blood pressure, bilirubin, and creatine kinase (OR ≈ 2.6), although the fully adjusted model was restricted to patients with complete data on all covariates.

## Odds Ratio

```{r}
exp(cbind(
  OR = coef(model_main),
  confint(model_main)
))
```
A one-unit increase in log-transformed lactate was associated with a 2.82-fold increase in the odds of ICU mortality (95% CI 2.66–2.99).

No subgroup analyses were performed due to the observational nature of the study and the absence of pre-specified effect modification hypotheses.

***

\newpage

## Inference and Statistical Analysis

### Hypothesis testing: Lactate versus Mortality

We applying the non parametric test called Wilconox test, which performs one- 
and two-sample Wilcoxon tests on vectors of data. The latter is also known 
as ‘Mann-Whitney’ test.

```{r}
# Split in survivors and non-survivors vectors
survivors <- analysis_df %>% filter(icu_dead == 0) %>% pull(log_lactate)
non_survivors <- analysis_df %>% filter(icu_dead == 1) %>% pull(log_lactate)
```

We procede to run a non-parametric test based on two hypothesis:

* $H_0$: Null hypothesis, which the lactate levels are not different between survivors and non-survivors
* $H_1$: Alternative hypothesis, which the lactate levels differ between survivors and non-survivors

```{r}
# Mann-Whitney U Test
wilcox_result <- wilcox.test(log_lactate ~ icu_dead, 
                              data = analysis_df,
                              conf.int = TRUE)

```


```{r}
cat(sprintf("Median lactate survivors: %.2f mmol/L [IQR: %.2f-%.2f]\n",
            median(survivors), 
            quantile(survivors, 0.25), 
            quantile(survivors, 0.75)))
cat(sprintf(" Median lactate non-survivors: %.2f mmol/L [IQR: %.2f-%.2f]\n",
            median(non_survivors), 
            quantile(non_survivors, 0.25), 
            quantile(non_survivors, 0.75)))
```


```{r}
# Show the results
wilcox_result
```
Based in the first hypothesis the median lactate levels were lower in survivors 
than in non-survivors:

* Survivors: 0.53 mmol/L (IQR 0.18–0.99)
* Non-survivors: 0.92 mmol/L (IQR 0.41–1.55)

This indicates a right-shifted distribution of lactate values among non-survivors.

After the no parametric Wilcox test, was used to compare lactate levels between
survivors and non-survivors. Because the analysis was performed on log-transformed
lactate values, the test evaluates differences in the location (median) of 
log-lactate between groups.

* W = 23,000,966
* p < 2.2 × 10⁻¹⁶
the Estimated location shift (log scale): −0.38
the 95% CI: −0.41 to −0.36

The p-value is low and indicates a highly statistically significant difference 
in lactate levels between survivors and non-survivors.

The negative location shift means that survivors have lower log-lactate 
values than non-survivors.

The confidence interval does not cross zero, reinforcing the robustness of the finding.

Because the test was conducted on log-lactate, the estimated difference 
corresponds to a multiplicative effect on the original scale and suggests that 
median lactate levels in survivors are approximately 30–35% lower than in non-survivors.

***

### Dose-response Analysis: Cochran-Armitage trend test.

The Cochran-Armitage trend test is a statistical method for detecting a linear 
trend in proportions across ordered categories, often used in dose-response 
studies or genetic association studies to see if a response 
(like disease presence) changes consistently (increases or decreases) 
with ordered exposures (like drug dose or genotype). 

It's a powerful modification of the chi-squared test, focusing specifically on 
the linear relationship, making it sensitive to gradual changes rather than
just overall differences.

```{r}
# Create lactate categories
analysis_df <- analysis_df %>%
  mutate(
    lactate_cat = case_when(
      lactate_value_num < 2 ~ "Normal",
      lactate_value_num >= 2 & lactate_value_num < 4 ~ "Elevated",
      lactate_value_num >= 4 ~ "Severe"
    ) %>% factor(levels = c("Normal", "Elevated", "Severe"))
  )

# Contingency table
cont_table <- table(analysis_df$lactate_cat, analysis_df$icu_dead)

cat("\nMortality by Lactate Category:\n")
mort_by_cat <- analysis_df %>%
  group_by(lactate_cat) %>%
  summarise(
    n = n(),
    deaths = sum(icu_dead),
    mortality_rate = mean(icu_dead) * 100,
    .groups = "drop"
  )
print(mort_by_cat)
```

```{r}
# Chi-square test
chisq_result <- chisq.test(cont_table)
cat(sprintf("\nChi-square test: χ² = %.2f, df = %d, p < 0.001\n",
            chisq_result$statistic, chisq_result$parameter))

# Cochran-Armitage trend test
library(DescTools)
ca_test <- CochranArmitageTest(cont_table, alternative = "one.sided")
cat(sprintf("Cochran-Armitage trend test: Z = %.2f, p = %.3f\n", 
            ca_test$statistic, ca_test$p.value))

# Interpretation based on the direction
if (ca_test$statistic > 0) {
  cat("Interpretation: Significant increasing trend in mortality with lactate\n")
} else {
  cat("Interpretation: Significant decreasing trend in mortality with lactate\n")
}
```
The statistical analyses demonstrate a highly significant association between 
lactate levels and ICU mortality.

The chi-square test yielded $\chi^2 = 1259.55$ (df = 2, p < 0.001),
indicating a strong relationship between lactate categories and death, while the
Cochran-Armitage trend test (Z = -31.50, p < 0.001) confirms an extremely 
strong linear trend.

Despite the negative Z statistic (an artifact of table coding), the mortality 
data unambiguously show an increasing dose-response relationship: patients with
normal lactate (<0.5 mmol/L) had 11.2% mortality, those with elevated 
lactate (0.5-2.0 mmol/L) had 15.2% mortality, and those with 
severe hyperlactatemia (>2.0 mmol/L) had 38.0% mortality, representing a more 
than three-fold increase in risk compared to the normal group. 

These findings confirm that higher lactate levels are strongly associated with
progressively worse outcomes in critically ill patients, establishing lactate 
as a powerful prognostic marker in the ICU setting.

***

### Confidence intervals: Mortality rates by category


```{r}

# Function for exact binomial CI
calculate_mortality_ci <- function(data, category) {
  subset_data <- data %>% filter(lactate_cat == category)
  n <- nrow(subset_data)
  deaths <- sum(subset_data$icu_dead)
  
  # Wilson score interval (better for proportions)
  binom_test <- binom.test(deaths, n)
  
  data.frame(
    category = category,
    n = n,
    deaths = deaths,
    mortality_rate = deaths/n * 100,
    ci_lower = binom_test$conf.int[1] * 100,
    ci_upper = binom_test$conf.int[2] * 100
  )
}

mortality_ci <- bind_rows(
  calculate_mortality_ci(analysis_df, "Normal"),
  calculate_mortality_ci(analysis_df, "Elevated"),
  calculate_mortality_ci(analysis_df, "Severe")
)

cat("\nMortality Rates with 95% Confidence Intervals:\n")
print(mortality_ci, digits = 2)
```

The 95% confidence intervals for mortality rates demonstrate precise estimates 
with non-overlapping ranges across lactate categories, confirming distinct risk strata. 

For patients with normal lactate levels, the mortality rate of 11.2% has a 
relatively narrow confidence interval (10.6% to 11.7%), reflecting the large 
sample size of 12,093 patients. The elevated lactate group shows a mortality 
rate of 15.2% (95% CI: 14.4% to 16.1%), with the lower bound of this interval 
clearly exceeding the upper bound of the normal group, indicating a statistically 
significant difference between these categories. 

Most strikingly, the severe hyperlactatemia group demonstrates a mortality 
rate of 38.0% (95% CI: 36.2% to 39.8%), with its lower confidence bound substantially higher than the upper bound of the elevated group, further confirming the progressive increase in risk. 

The non-overlapping confidence intervals provide strong statistical evidence that each lactate category represents a genuinely distinct risk level, with patients moving from one category to the next facing significantly and progressively higher mortality risk.

***

### Confidence intervals using bootstrap for odds ratios.


```{r}
# Function to calculate OR via bootstrap
boot_or <- function(data, indices) {
  d <- data[indices, ]
  model <- glm(icu_dead ~ log_lactate + age_icu_exit + gender,
               data = d, family = binomial())
  coef(model)["log_lactate"]
}

# Run bootstrap (1000 replicates)
set.seed(42)
boot_results <- boot(data = analysis_df, statistic = boot_or, R = 1000)

# Calculate percentile CI
boot_ci <- boot.ci(boot_results, type = "perc")

cat("\nBootstrap OR for log-lactate (1000 replicates):\n")
cat(sprintf("   Point estimate: %.3f (OR = %.3f)\n", 
            boot_results$t0, 
            exp(boot_results$t0)))
cat(sprintf("   95%% Percentile CI: %.3f to %.3f\n",
            boot_ci$percent[4], 
            boot_ci$percent[5]))
cat(sprintf("   OR 95%% CI: %.3f to %.3f\n",
            exp(boot_ci$percent[4]), 
            exp(boot_ci$percent[5])))
```
The bootstrap analysis with 1,000 replicates confirms a robust association between lactate levels and mortality, with each unit increase in log-lactate associated with an odds ratio of 2.82 (95% CI: 2.66 to 3.01), meaning that for every doubling of lactate concentration, the odds of death increase approximately 2.8-fold. The confidence interval is relatively narrow and does not include 1.0, providing strong statistical evidence that elevated lactate is a significant and reliable predictor of ICU mortality, with the lower bound still indicating at least a 2.66-fold increased risk per unit change in log-lactate.

***

### Central Limit theorem demonstration

```{r}
# Demonstrate CLT by repeated sampling
set.seed(42)
n_simulations <- 10000
sample_size <- 100

sample_means <- replicate(n_simulations, {
  sample_idx <- sample(1:nrow(eda_df), sample_size, replace = TRUE)
  mean(eda_df$lactate_value_num[sample_idx])
})

# Test normality of sampling distribution
shapiro_test <- shapiro.test(sample(sample_means, 5000))

cat(sprintf("\nSampling distribution of lactate mean (%d samples, n=%d each):\n",
            n_simulations, sample_size))
cat(sprintf("   Mean of sample means: %.3f\n", mean(sample_means)))
cat(sprintf("   Population mean: %.3f\n", mean(eda_df$lactate_value_num)))
cat(sprintf("   SD of sample means: %.3f\n", sd(sample_means)))
cat(sprintf("   Theoretical SE: %.3f\n", 
            sd(eda_df$lactate_value_num) / sqrt(sample_size)))
cat(sprintf("   Shapiro-Wilk test: W = %.4f, p = %.4f\n",
            shapiro_test$statistic, shapiro_test$p.value))
cat("   Conclusion: Sampling distribution is approximately normal (CLT)\n")


fig10a <- ggplot(eda_df, aes(lactate_value_num)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(
    title = "Original Lactate Distribution",
    x = "Lactate (mmol/L)",
    y = "Count"
  )

fig10b <- ggplot(data.frame(sample_means), aes(sample_means)) +
  geom_histogram(bins = 50, fill = "coral", color = "white") +
  stat_function(
    fun = function(x)
      dnorm(x, mean(sample_means), sd(sample_means)) *
      length(sample_means) *
      diff(range(sample_means)) / 50,
    color = "red",
    linewidth = 1
  ) +
  labs(
    title = "Sampling Distribution of Mean",
    x = "Sample Mean Lactate (mmol/L)",
    y = "Count"
  )

fig10a
fig10b

```



```{r}
ggsave(
  filename = "../figures/figure10a_clt_demonstration.png",
  plot = fig10a,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)

ggsave(
  filename = "../figures/figure10b_clt_demonstration.png",
  plot = fig10b,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```


The central limit theorem demonstration using 10,000 bootstrap samples of size 
100 confirms that the sampling distribution of lactate means is approximately normal. 
The mean of all sample means (2.438 mmol/L) is virtually identical to the 
population mean (2.436 mmol/L), demonstrating unbiased estimation. 
The observed standard deviation of sample means (0.206) closely matches the 
theoretical standard error (0.205), validating the mathematical prediction of
sampling variability. Although the Shapiro-Wilk test shows statistical 
significance (W = 0.9944, p < 0.001) due to the large number of samples making 
it highly sensitive to minor deviations, the W statistic of 0.9944 
(very close to the perfect value of 1.0) indicates the distribution is 
substantively normal, confirming that the central limit theorem holds even for 
lactate data which may not be normally distributed at the individual level.

*** 

### Probability estimation: Predicted mortality probabilities

```{r}
# Fit final model
final_model <- glm(icu_dead ~ log_lactate + age_icu_exit + gender,
                   data = analysis_df, family = binomial())

# Create prediction scenarios
scenarios <- data.frame(
  scenario = c("Low risk", "Medium risk", "High risk", "Very high risk"),
  log_lactate = log(c(1.0, 2.0, 4.0, 8.0)),
  age_icu_exit = c(50, 65, 75, 80),
  gender = c("M", "M", "M", "M")
)

# Predict probabilities
scenarios$predicted_prob <- predict(final_model, scenarios, type = "response")
scenarios$predicted_mortality_pct <- scenarios$predicted_prob * 100

# Calculate 95% CI for predictions
pred_se <- predict(final_model, scenarios, type = "link", se.fit = TRUE)
scenarios$ci_lower <- plogis(pred_se$fit - 1.96 * pred_se$se.fit) * 100
scenarios$ci_upper <- plogis(pred_se$fit + 1.96 * pred_se$se.fit) * 100

cat("\nPredicted Mortality for Clinical Scenarios:\n")
print(scenarios %>% 
        select(scenario, predicted_mortality_pct, ci_lower, ci_upper) %>%
        mutate(across(where(is.numeric), ~round(., 1))),
      row.names = FALSE)
```

The logistic regression model provides well-calibrated probability estimates for mortality across different risk scenarios with tight confidence intervals. Patients in the low-risk category have a predicted mortality of 5.3% (95% CI: 4.9% to 5.8%), representing the baseline risk for those with favorable lactate levels. Medium-risk patients face a predicted mortality of 14.1% (95% CI: 13.4% to 14.7%), nearly tripling the baseline risk, while high-risk patients show a predicted mortality of 29.8% (95% CI: 28.5% to 31.0%), approaching one-third mortality probability. Most critically, very high-risk patients have a predicted mortality of 49.4% (95% CI: 47.1% to 51.7%), indicating nearly a coin-flip chance of death and representing more than a nine-fold increase compared to low-risk patients. The narrow, non-overlapping confidence intervals across all risk categories demonstrate that the model provides reliable and discriminating probability estimates, enabling clinicians to stratify patients into distinct prognostic groups based on their lactate levels.

***

### Subgroup analysis: Age stratification

```{r}
# Create age groups
analysis_df <- analysis_df %>%
  mutate(
    age_group = cut(age_icu_exit, 
                    breaks = c(0, 50, 65, 80, Inf),
                    labels = c("<50", "50-65", "65-80", ">80"))
  )

# Stratified analysis
subgroup_results <- analysis_df %>%
  group_by(age_group) %>%
  do({
    model <- glm(icu_dead ~ log_lactate, data = ., family = binomial())
    tidy(model, exponentiate = TRUE, conf.int = TRUE) %>%
      filter(term == "log_lactate")
  }) %>%
  ungroup()

cat("\nOR for log-lactate by age group:\n")
print(subgroup_results %>% 
        select(age_group, estimate, conf.low, conf.high, p.value),
      digits = 3)
```


```{r}
# Test for interaction
model_main <- glm(icu_dead ~ log_lactate + age_group,
                  data = analysis_df, family = binomial())

model_interaction <- glm(icu_dead ~ log_lactate * age_group,
                         data = analysis_df, family = binomial())

anova_interaction <- anova(model_main, model_interaction, test = "Chisq")

cat(sprintf("\nInteraction test (lactate × age): χ² = %.2f, p = %.3f\n",
            anova_interaction$Deviance[2],
            anova_interaction$`Pr(>Chi)`[2]))

if (anova_interaction$`Pr(>Chi)`[2] > 0.05) {
  cat("Conclusion: No significant effect modification by age\n")
} else {
  cat("Conclusion: Significant effect modification by age detected\n")
}
```

The interaction test reveals significant effect modification by age (χ² = 49.52, p < 0.001), indicating that the association between lactate levels and mortality is not uniform across different age groups. This means that the prognostic value of lactate varies depending on patient age, with some age strata showing stronger or weaker associations between elevated lactate and death than others. The highly significant p-value provides strong statistical evidence that age modifies how lactate predicts mortality risk, suggesting that clinicians should interpret lactate values in the context of patient age when assessing prognosis, as a given lactate level may carry different mortality implications for younger versus older ICU patients.

### Statistical Inference analysis Summary

The statistical analysis demonstrates robust and significant associations between lactate levels and ICU mortality through multiple complementary approaches. Non-parametric testing confirmed that non-survivors had significantly higher lactate levels than survivors (Wilcoxon W = 23,000,966, p < 0.001), with a median difference of approximately 0.4 mmol/L on the log scale. Categorical analysis revealed a strong dose-response relationship (χ² = 1259.55, df = 2, p < 0.001) with mortality rates increasing from 11.2% in normal lactate patients to 38.0% in those with severe hyperlactatemia, confirmed by the Cochran-Armitage trend test (|Z| = 31.50, p < 0.001). Bootstrap analysis with 1,000 replicates established that each unit increase in log-lactate increases mortality odds by 2.82-fold (95% CI: 2.66-3.01), demonstrating excellent precision and reliability. The model generates well-calibrated probability estimates ranging from 5.3% mortality in low-risk patients to 49.4% in very high-risk patients, with narrow, non-overlapping confidence intervals confirming strong discrimination. Central limit theorem validation using 10,000 samples confirmed appropriate normality of the sampling distribution (W = 0.9944), supporting the validity of parametric inference. Importantly, subgroup analysis revealed significant effect modification by age (χ² = 49.52, p < 0.001), indicating that lactate's prognostic value varies across age strata and should be interpreted in clinical context.

***

\newpage

# Machine Learning modeling

## Purpose: Patient clustering and segmentation to identify risk phenotypes

This analysis aims to identify distinct patient subgroups or clinical phenotypes within the ICU population by clustering patients based on their lactate profiles and associated characteristics. 

By segmenting patients into homogeneous risk groups, we can uncover hidden patterns that may not be apparent through traditional statistical approaches, potentially revealing clinically meaningful patient subtypes with different underlying pathophysiology, treatment responses, or prognostic trajectories. 

This unsupervised learning approach allows for data-driven discovery of patient phenotypes that could inform personalized risk stratification, targeted interventions, and improved resource allocation in critical care settings.

```{r}
# Use complete cases for clustering (from your PDF: 141 patients with all variables)
clustering_df <- analysis_df %>%
  select(icustay_id, age_icu_exit, log_lactate, log_ck, 
         log_bilirubin, mabp_value_num, icu_dead) %>%
  na.omit() %>%
  mutate(
    patient_id = row_number()
  )

cat(sprintf("Clustering dataset: %d patients with complete data\n", 
            nrow(clustering_df)))
cat(sprintf("Mortality rate: %.1f%%\n\n", 
            mean(clustering_df$icu_dead) * 100))

# Prepare feature matrix (exclude IDs and outcome)
features_for_clustering <- clustering_df %>%
  select(age_icu_exit, log_lactate, log_ck, log_bilirubin, mabp_value_num)

# Standardize features (z-score normalization)
features_scaled <- scale(features_for_clustering)
rownames(features_scaled) <- clustering_df$patient_id
```

We proceed to determinate the optimal number of clusters.


```{r}
# a) Elbow Method
set.seed(42)
wss <- sapply(1:10, function(k) {
  kmeans(features_scaled, centers = k, nstart = 25)$tot.withinss
})

# b) Silhouette Method
set.seed(42)
silhouette_scores <- sapply(2:10, function(k) {
  km <- kmeans(features_scaled, centers = k, nstart = 25)
  ss <- silhouette(km$cluster, dist(features_scaled))
  mean(ss[, 3])
})

# c) Gap Statistic
set.seed(42)
gap_stat <- clusGap(features_scaled, 
                    FUN = kmeans, 
                    nstart = 25,
                    K.max = 10, 
                    B = 50)

cat("\nOptimal clusters by different methods:\n")
cat(sprintf("  Elbow method: Visual inspection of plot\n"))
cat(sprintf("  Silhouette: k = %d (score = %.3f)\n", 
            which.max(silhouette_scores) + 1, 
            max(silhouette_scores)))
cat(sprintf("  Gap statistic: k = %d\n", 
            which.max(gap_stat$Tab[, "gap"])))

# Plot determination methods
png("../figures/figure10_optimal_clusters.png",
    width = 15, height = 5, units = "in", res = 300)

par(mfrow = c(1, 3))

# Elbow plot
plot(1:10, wss, type = "b", pch = 19, col = "steelblue",
     xlab = "Number of Clusters (k)", 
     ylab = "Within-cluster Sum of Squares",
     main = "Elbow Method")
grid()

# Silhouette plot
plot(2:10, silhouette_scores, type = "b", pch = 19, col = "coral",
     xlab = "Number of Clusters (k)", 
     ylab = "Average Silhouette Width",
     main = "Silhouette Method")
grid()

# Gap statistic plot
plot(gap_stat, main = "Gap Statistic Method")

dev.off()

cat("\nFigure saved: results/figures/figure10_optimal_clusters.png\n")

# Choose k=3 based on clinical interpretability and silhouette
optimal_k <- 3
cat(sprintf("\nSelected k = %d for final clustering\n", optimal_k))
```

After the first results, we have a k=[2,3] to add on our clustering algorithm.

## Apply K-Means Clustering

```{r}
set.seed(42)
kmeans_result <- kmeans(features_scaled, centers = optimal_k, nstart = 50)

# Add cluster assignment to dataframe
clustering_df$cluster_kmeans <- factor(kmeans_result$cluster,
                                       labels = c("Cluster 1", "Cluster 2", "Cluster 3"))

# Cluster sizes
cat("\nCluster sizes:\n")
print(table(clustering_df$cluster_kmeans))

# Cluster centers (back-transformed)
cluster_centers <- kmeans_result$centers
cluster_centers_original <- t(apply(cluster_centers, 1, function(x) {
  x * attr(features_scaled, "scaled:scale") + 
    attr(features_scaled, "scaled:center")
}))

cat("\nCluster centers (original scale):\n")
print(round(cluster_centers_original, 2))

# Mortality by cluster
mortality_by_cluster <- clustering_df %>%
  group_by(cluster_kmeans) %>%
  summarise(
    n = n(),
    mean_age = mean(age_icu_exit),
    mean_lactate = mean(exp(log_lactate)),  # Back-transform
    mean_ck = mean(exp(log_ck)),
    mean_bilirubin = mean(exp(log_bilirubin)),
    mean_mabp = mean(mabp_value_num),
    deaths = sum(icu_dead),
    mortality_rate = mean(icu_dead) * 100,
    .groups = "drop"
  )

cat("\nMortality by K-means cluster:\n")
print(mortality_by_cluster, digits = 1)

# Statistical test: Chi-square
chisq_kmeans <- chisq.test(table(clustering_df$cluster_kmeans, 
                                  clustering_df$icu_dead))
cat(sprintf("\nChi-square test (cluster vs mortality): χ² = %.2f, p = %.4f\n",
            chisq_kmeans$statistic, chisq_kmeans$p.value))
```
The k-means clustering analysis identified three distinct patient phenotypes with significantly different mortality profiles (χ² = 79.67, p < 0.001). Cluster 1 represents the largest group (591 patients) characterized by older age (70 years), low lactate levels, moderate creatine kinase, low bilirubin, and relatively stable blood pressure, suggesting a lower-risk elderly population. Cluster 2 (314 patients) captures high-risk patients with elevated lactate (1.41 log mmol/L), very high bilirubin (1.04 log mg/dL), elevated creatine kinase, and lower blood pressure (73.71 mmHg), indicating multi-organ dysfunction and hemodynamic instability. Cluster 3 (239 patients) identifies a younger cohort (50 years) with moderately elevated lactate, extremely high creatine kinase (7.41 log U/L), and good blood pressure, potentially representing trauma or rhabdomyolysis cases. These data-driven phenotypes reveal that ICU patients stratify into clinically interpretable risk groups beyond simple lactate categorization, with each cluster representing distinct pathophysiological patterns that significantly associate with mortality outcomes.

# ============================================================================
# 3. HIERARCHICAL CLUSTERING
# ============================================================================

```{r}
# Calculate distance matrix
dist_matrix <- dist(features_scaled, method = "euclidean")

# Perform hierarchical clustering
hc_complete <- hclust(dist_matrix, method = "complete")
hc_ward <- hclust(dist_matrix, method = "ward.D2")

# Cut dendrogram to get 3 clusters
clustering_df$cluster_hc <- factor(cutree(hc_ward, k = optimal_k),
                                   labels = c("HC-1", "HC-2", "HC-3"))

# Mortality by hierarchical cluster
mortality_by_hc <- clustering_df %>%
  group_by(cluster_hc) %>%
  summarise(
    n = n(),
    deaths = sum(icu_dead),
    mortality_rate = mean(icu_dead) * 100,
    .groups = "drop"
  )

cat("\nMortality by Hierarchical cluster:\n")
print(mortality_by_hc, digits = 1)

# Plot dendrogram
png("../figures/figure11_dendrogram.png",
    width = 12, height = 6, units = "in", res = 300)

plot(hc_ward, labels = FALSE, hang = -1,
     main = "Hierarchical Clustering Dendrogram (Ward's Method)",
     xlab = "Patients", ylab = "Height")
rect.hclust(hc_ward, k = optimal_k, border = 2:4)
```
The hierarchical clustering analysis identified three patient phenotypes with distinct mortality profiles that differ substantially from the k-means results. Cluster HC-1, the largest group with 371 patients, shows a relatively low mortality rate of 18.9%, representing a moderate-risk population. Cluster HC-2, comprising 593 patients, demonstrates the highest mortality at 38.3%, indicating a high-risk phenotype characterized by severe physiological derangement and poor prognosis. Cluster HC-3, with 180 patients, exhibits an intermediate mortality rate of 21.7%, falling between the other two clusters. The hierarchical approach reveals a different patient segmentation pattern compared to k-means, with HC-2 identifying a larger proportion of critically ill patients at substantially elevated mortality risk, suggesting that the dendrogram-based hierarchical method captures different aspects of patient heterogeneity and risk stratification that may complement the k-means partitioning approach.

### Model-based clustering (Gaussian Mixture models)


```{r}
# Fit GMM with 1-6 components
set.seed(42)
gmm_result <- Mclust(features_scaled, G = 1:6)

cat(sprintf("\nOptimal GMM model: %s with %d components\n",
            gmm_result$modelName, gmm_result$G))
cat(sprintf("BIC: %.1f\n", gmm_result$bic))

# Add GMM cluster assignment
clustering_df$cluster_gmm <- factor(gmm_result$classification,
                                    labels = paste0("GMM-", 1:gmm_result$G))

# Mortality by GMM cluster
mortality_by_gmm <- clustering_df %>%
  group_by(cluster_gmm) %>%
  summarise(
    n = n(),
    deaths = sum(icu_dead),
    mortality_rate = mean(icu_dead) * 100,
    .groups = "drop"
  )

cat("\nMortality by GMM cluster:\n")
print(mortality_by_gmm, digits = 1)
```


### Cluster Visualization PCA


```{r}
# Perform PCA for visualization
pca_result <- prcomp(features_scaled)
pca_variance <- summary(pca_result)$importance[2, 1:2] * 100

cat(sprintf("\nPCA variance explained:\n"))
cat(sprintf("  PC1: %.1f%%\n", pca_variance[1]))
cat(sprintf("  PC2: %.1f%%\n", pca_variance[2]))

# Add PCA coordinates to dataframe
clustering_df$PC1 <- pca_result$x[, 1]
clustering_df$PC2 <- pca_result$x[, 2]

# Create comprehensive visualization
png("../figures/figure12_cluster_visualization.png",
    width = 15, height = 10, units = "in", res = 300)

par(mfrow = c(2, 2))

# K-means clusters
plot(clustering_df$PC1, clustering_df$PC2, 
     col = as.numeric(clustering_df$cluster_kmeans) + 1,
     pch = ifelse(clustering_df$icu_dead == 1, 17, 19),
     cex = 1.2,
     xlab = sprintf("PC1 (%.1f%% variance)", pca_variance[1]),
     ylab = sprintf("PC2 (%.1f%% variance)", pca_variance[2]),
     main = "K-means Clustering")
legend("topright", 
       legend = c(levels(clustering_df$cluster_kmeans), "Died", "Survived"),
       col = c(2:4, 1, 1),
       pch = c(19, 19, 19, 17, 19),
       cex = 0.8)

# Hierarchical clusters
plot(clustering_df$PC1, clustering_df$PC2, 
     col = as.numeric(clustering_df$cluster_hc) + 1,
     pch = ifelse(clustering_df$icu_dead == 1, 17, 19),
     cex = 1.2,
     xlab = sprintf("PC1 (%.1f%% variance)", pca_variance[1]),
     ylab = sprintf("PC2 (%.1f%% variance)", pca_variance[2]),
     main = "Hierarchical Clustering")
legend("topright", 
       legend = levels(clustering_df$cluster_hc),
       col = 2:4, pch = 19, cex = 0.8)

# GMM clusters
plot(clustering_df$PC1, clustering_df$PC2, 
     col = as.numeric(clustering_df$cluster_gmm) + 1,
     pch = ifelse(clustering_df$icu_dead == 1, 17, 19),
     cex = 1.2,
     xlab = sprintf("PC1 (%.1f%% variance)", pca_variance[1]),
     ylab = sprintf("PC2 (%.1f%% variance)", pca_variance[2]),
     main = "GMM Clustering")
legend("topright", 
       legend = levels(clustering_df$cluster_gmm),
       col = 2:(gmm_result$G + 1), pch = 19, cex = 0.8)

# Mortality overlay
plot(clustering_df$PC1, clustering_df$PC2, 
     col = ifelse(clustering_df$icu_dead == 1, "red", "green"),
     pch = 19, cex = 1.2,
     xlab = sprintf("PC1 (%.1f%% variance)", pca_variance[1]),
     ylab = sprintf("PC2 (%.1f%% variance)", pca_variance[2]),
     main = "Mortality Status")
legend("topright", 
       legend = c("Died", "Survived"),
       col = c("red", "green"), pch = 19, cex = 0.8)
```
```{r}
summary(pca_result)$importance[,] * 100
```
The first principal component has only ~30% of variance explained, plus the second
component is around 54%. We need almost 4 principal components to reach 90%.


### Cluster profiling & interpretation 

```{r}
# Detailed profiling
cluster_profile <- clustering_df %>%
  group_by(cluster_kmeans) %>%
  summarise(
    n = n(),
    
    # Demographics
    age_mean = mean(age_icu_exit),
    age_sd = sd(age_icu_exit),
    
    # Lactate (back-transformed)
    lactate_median = median(exp(log_lactate)),
    lactate_q1 = quantile(exp(log_lactate), 0.25),
    lactate_q3 = quantile(exp(log_lactate), 0.75),
    
    # CK (back-transformed)
    ck_median = median(exp(log_ck)),
    ck_q1 = quantile(exp(log_ck), 0.25),
    ck_q3 = quantile(exp(log_ck), 0.75),
    
    # Bilirubin (back-transformed)
    bili_median = median(exp(log_bilirubin)),
    bili_q1 = quantile(exp(log_bilirubin), 0.25),
    bili_q3 = quantile(exp(log_bilirubin), 0.75),
    
    # MAP
    map_mean = mean(mabp_value_num),
    map_sd = sd(mabp_value_num),
    
    # Outcome
    deaths = sum(icu_dead),
    mortality_pct = mean(icu_dead) * 100,
    
    .groups = "drop"
  )
```



```{r}
cat("\nDetailed Cluster Profiles:\n")
print(cluster_profile, digits = 2)

# Clinical interpretation
cat("\n\nCLINICAL INTERPRETATION OF CLUSTERS:\n")
cat(rep("=", 70), "\n")

for (i in 1:optimal_k) {
  cluster_data <- cluster_profile %>% filter(cluster_kmeans == paste0("Cluster ", i))
  
  cat(sprintf("\n%s (n=%d, mortality %.1f%%):\n", 
              paste0("Cluster ", i), 
              cluster_data$n, 
              cluster_data$mortality_pct))
  
  cat(sprintf("  Age: %.0f ± %.0f years\n", 
              cluster_data$age_mean, 
              cluster_data$age_sd))
  cat(sprintf("  Lactate: %.1f [%.1f-%.1f] mmol/L\n",
              cluster_data$lactate_median,
              cluster_data$lactate_q1,
              cluster_data$lactate_q3))
  cat(sprintf("  CK: %.0f [%.0f-%.0f] IU/L\n",
              cluster_data$ck_median,
              cluster_data$ck_q1,
              cluster_data$ck_q3))
  cat(sprintf("  Bilirubin: %.1f [%.1f-%.1f] mg/dL\n",
              cluster_data$bili_median,
              cluster_data$bili_q1,
              cluster_data$bili_q3))
  cat(sprintf("  MAP: %.0f ± %.0f mmHg\n",
              cluster_data$map_mean,
              cluster_data$map_sd))
  
  # Interpret cluster
  if (cluster_data$lactate_median < 2 && cluster_data$mortality_pct < 30) {
    cat("  INTERPRETATION: LOW RISK - Relatively stable patients\n")
  } else if (cluster_data$lactate_median >= 4 || cluster_data$mortality_pct > 60) {
    cat("  INTERPRETATION: HIGH RISK - Severely ill, high mortality\n")
  } else {
    cat("  INTERPRETATION: MODERATE RISK - Mixed severity\n")
  }
}
```

The cluster profiling reveals three clinically distinct ICU patient phenotypes with different risk profiles and pathophysiological characteristics. Cluster 1 (n=591, 20.8% mortality) represents a "low-risk elderly stable" group with older patients (70 years) who have relatively normal lactate (1.4 mmol/L), low creatine kinase, stable hemodynamics (MAP 82 mmHg), and preserved liver function, suggesting age-related admissions with less acute physiological derangement. Cluster 2 (n=314, 48.7% mortality) identifies a "high-risk multi-organ dysfunction" phenotype characterized by severe hyperlactatemia (4.1 mmol/L), significantly elevated bilirubin (2.5 mg/dL), hemodynamic instability (MAP 74 mmHg), and elevated creatine kinase, indicating patients with profound metabolic derangement, hepatic dysfunction, and tissue hypoperfusion who face nearly 50% mortality risk. Cluster 3 (n=239, 25.1% mortality) captures a "moderate-risk younger trauma/rhabdomyolysis" group with substantially younger patients (50 years), extremely elevated creatine kinase (1,465 IU/L suggesting muscle injury), moderately elevated lactate, preserved blood pressure, and normal bilirubin, likely representing trauma, seizure, or rhabdomyolysis cases with intermediate prognosis despite marked CK elevation, demonstrating that younger age and single-organ pathology confer better outcomes than multi-organ failure.


```{r}
# Save clustering results
clustering_results <- list(
  optimal_k = optimal_k,
  kmeans_model = kmeans_result,
  hc_model = hc_ward,
  gmm_model = gmm_result,
  mortality_by_cluster = mortality_by_cluster,
  cluster_profiles = cluster_profile,
  patient_assignments = clustering_df
)

if (!dir.exists("../results")) {
dir.create("../results", recursive = TRUE)
}
save(clustering_results, file = "../results/clustering_results.RData")

cat("\nKey findings:\n")
cat(sprintf("  - Identified %d distinct patient clusters\n", optimal_k))
cat(sprintf("  - Mortality ranges from %.1f%% to %.1f%% across clusters\n",
            min(mortality_by_cluster$mortality_rate),
            max(mortality_by_cluster$mortality_rate)))
cat("  - Clusters reflect different severity profiles\n")
cat("\nResults saved to: results/clustering_results.RData\n")
```

***

## ML models (C5.0, Random Forest, XGBoost) + SHAP explanations


```{r}
# Prepare ML dataset with median imputation
ml_data <- analysis_df %>%
  select(icustay_id, age_icu_exit, gender, icu_dead, 
         log_lactate, log_ck, log_bilirubin, mabp_value_num) %>%
  mutate(
    # Impute missing values with median
    log_ck = if_else(is.na(log_ck), median(log_ck, na.rm = TRUE), log_ck),
    log_bilirubin = if_else(is.na(log_bilirubin), 
                             median(log_bilirubin, na.rm = TRUE), 
                             log_bilirubin),
    mabp_value_num = if_else(is.na(mabp_value_num), 
                              median(mabp_value_num, na.rm = TRUE), 
                              mabp_value_num),
    # Convert gender to numeric
    gender_num = ifelse(gender == "M", 1, 0),
    # Convert outcome to factor for classification
    icu_dead_f = factor(icu_dead, levels = c(0, 1), 
                        labels = c("Survived", "Died"))
  ) %>%
  select(-gender, -icustay_id)

cat(sprintf("Dataset size: %d patients\n", nrow(ml_data)))
cat(sprintf("Features: %d\n", ncol(ml_data) - 2))  # Exclude outcomes
cat(sprintf("Class balance: %.1f%% died, %.1f%% survived\n",
            mean(ml_data$icu_dead) * 100,
            mean(ml_data$icu_dead == 0) * 100))

# Train-test split (75-25)
set.seed(42)
train_idx <- createDataPartition(ml_data$icu_dead_f, p = 0.75, list = FALSE)
train_data <- ml_data[train_idx, ]
test_data <- ml_data[-train_idx, ]

cat(sprintf("\nTrain set: %d (%.1f%% died)\n", 
            nrow(train_data),
            mean(train_data$icu_dead) * 100))
cat(sprintf("Test set: %d (%.1f%% died)\n", 
            nrow(test_data),
            mean(test_data$icu_dead) * 100))
```
Next step is creating a machine learning baseline model, the Logistic regression.

```{r}
# Train control
train_control <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# Train logistic regression
set.seed(42)
model_glm <- train(
  icu_dead_f ~ age_icu_exit + gender_num + log_lactate + 
    log_ck + log_bilirubin + mabp_value_num,
  data = train_data,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "ROC"
)

# Predictions
pred_glm <- predict(model_glm, test_data, type = "prob")[, "Died"]
roc_glm <- roc(test_data$icu_dead_f, pred_glm)

cat(sprintf("Logistic Regression AUC: %.3f (%.3f-%.3f)\n",
            auc(roc_glm), ci.auc(roc_glm)[1], ci.auc(roc_glm)[3]))
```

The first results as baseline is Area Under Curve around 70%.


### DECISION TREE: C5.0 ALGORITHM

```{r}
# Train C5.0 decision tree
set.seed(42)
model_c50 <- C5.0(
  icu_dead_f ~ age_icu_exit + gender_num + log_lactate + 
    log_ck + log_bilirubin + mabp_value_num,
  data = train_data,
  rules = FALSE,  # Tree format
  trials = 1      # Single tree (no boosting)
)

# Print tree summary
cat("\nC5.0 Tree Summary:\n")
summary(model_c50)
```

```{r}
# Extract rules
cat("\n\nDECISION RULES FROM C5.0:\n")
cat(rep("=", 70), "\n")
model_c50_rules <- C5.0(
  icu_dead_f ~ age_icu_exit + gender_num + log_lactate + 
    log_ck + log_bilirubin + mabp_value_num,
  data = train_data,
  rules = TRUE,  # Rule format
  trials = 1
)

cat("\n")
summary(model_c50_rules)

# Predictions
pred_c50 <- predict(model_c50, test_data, type = "prob")[, "Died"]
roc_c50 <- roc(test_data$icu_dead_f, pred_c50)

cat(sprintf("\nC5.0 Decision Tree AUC: %.3f (%.3f-%.3f)\n",
            auc(roc_c50), ci.auc(roc_c50)[1], ci.auc(roc_c50)[3]))

# Variable importance
c50_importance <- C5imp(model_c50, metric = "usage")
cat("\nC5.0 Variable Importance (usage-based):\n")
print(c50_importance)
```

```{r}
plot(model_c50)
```


```{r}
summary(model_c50_rules)
```

The C5.0 decision tree generated six interpretable rules that classify ICU patients 
into survival or death outcomes based on lactate, age, and bilirubin levels. 
Rules 1-3 identify survival patterns: Rule 1 predicts survival for younger 
patients (≤76 years) with moderate lactate levels (≤2.26 log mmol/L or ~9.6 mmol/L), 
covering 11,902 cases with 87.7% accuracy; Rule 2 predicts survival for all patients 
with low lactate (≤1.36 log mmol/L or ~3.9 mmol/L), the most protective factor 
covering 14,540 cases with 87.3% accuracy; and Rule 3 combines moderate 
lactate (≤1.63 log mmol/L or ~5.1 mmol/L) with normal bilirubin 
(≤1.93 log mg/dL or ~6.9 mg/dL) to predict survival in 15,090 cases with 87.3% accuracy. 

Rules 4-6 identify high-risk mortality patterns: Rule 4 shows that severe hyperlactatemia (>2.26 log mmol/L or >9.6 mmol/L) alone predicts death with 71.3% accuracy and a 4.5-fold lift; Rule 5 identifies elderly patients (>76 years) with elevated lactate (>1.63 log mmol/L or >5.1 mmol/L) as high-risk with 63.0% accuracy and 3.9-fold lift; and Rule 6 reveals that combined elevated lactate (>1.36 log mmol/L or >3.9 mmol/L) with high bilirubin (>1.93 log mg/dL or >6.9 mg/dL) predicts death with 59.9% accuracy and 3.7-fold lift, indicating multi-organ dysfunction. The model achieves 85% overall accuracy on training data, with lactate being the most critical variable (used 100% of the time), followed by bilirubin (91%) and age (73%), demonstrating that lactate thresholds around 4-10 mmol/L are key decision points, with elderly age and hepatic dysfunction significantly modifying mortality risk.




***

### 4. RANDOM FOREST

```{r}
# Train Random Forest
set.seed(42)
model_rf <- randomForest(
  icu_dead_f ~ age_icu_exit + gender_num + log_lactate + 
    log_ck + log_bilirubin + mabp_value_num,
  data = train_data,
  ntree = 500,
  mtry = 3,  # sqrt(6) ≈ 2.5
  importance = TRUE,
  nodesize = 5
)

cat("\nRandom Forest Model:\n")
print(model_rf)

# Predictions
pred_rf <- predict(model_rf, test_data, type = "prob")[, "Died"]
roc_rf <- roc(test_data$icu_dead_f, pred_rf)

cat(sprintf("\nRandom Forest AUC: %.3f (%.3f-%.3f)\n",
            auc(roc_rf), ci.auc(roc_rf)[1], ci.auc(roc_rf)[3]))

# Variable importance
cat("\nRandom Forest Variable Importance:\n")
rf_importance <- importance(model_rf)
print(rf_importance)

# Plot importance
png("../figures/figure13_rf_importance.png",
    width = 10, height = 7, units = "in", res = 300)

varImpPlot(model_rf, 
           main = "Random Forest Variable Importance",
           pch = 19,
           col = "steelblue",
           cex = 1.2)

dev.off()

cat("\nFigure saved: results/figures/figure13_rf_importance.png\n")
```


### 4. LIGHT GBM

```{r}
# Train LightGBM

train_matrix <- model.matrix(icu_dead_f ~ . - icu_dead - 1, data = train_data)
test_matrix <- model.matrix(icu_dead_f ~ . - icu_dead - 1, data = test_data)

train_labels <- as.numeric(train_data$icu_dead_f == "Died")
test_labels <- as.numeric(test_data$icu_dead_f == "Died")

dtest <- xgb.DMatrix(data = test_matrix, label = test_labels)

dtrain <- lgb.Dataset(
  data = train_matrix,
  label = train_labels
)

dtrain <- lgb.Dataset(
  data = test_matrix,
  label = test_labels
)

# LGBoost parameters
params <- list(
  objective = "binary",
  metric = "auc",
  learning_rate = 0.05,
  num_leaves = 15,      # similar complexity to C5.0
  max_depth = 4,
  min_data_in_leaf = 20,
  feature_fraction = 0.9,
  bagging_fraction = 0.8,
  bagging_freq = 5
)


# Training
set.seed(42)

lgb_model <- lgb.train(
  params = params,
  data = dtrain,
  nrounds = 300,
  verbose = -1
)


# Predictions
pred_prob <- predict(lgb_model, test_matrix)

roc_obj <- pROC::roc(test_labels, pred_prob)
cat(sprintf("Test AUC: %.3f\n", pROC::auc(roc_obj)))

```


```{r}
# Feature importance
importance_lgb <- lgb.importance(model = lgb_model)
cat("\nLightGBoost Feature Importance:\n")
print(importance_lgb)

# Plot importance
png("../figures/figure14_lgb_importance.png",
    width = 10, height = 7, units = "in", res = 300)
lgb.plot.importance(importance_lgb)

```


### Model Comparison

```{r}
# Create comparison table
model_comparison <- data.frame(
  Model = c("Logistic Regression", "C5.0 Decision Tree", 
            "Random Forest", "LightGBoost"),
  AUC = c(auc(roc_glm), auc(roc_c50), auc(roc_rf), auc(roc_obj)),
  CI_lower = c(ci.auc(roc_glm)[1], ci.auc(roc_c50)[1], 
               ci.auc(roc_rf)[1], ci.auc(roc_obj)[1]),
  CI_upper = c(ci.auc(roc_glm)[3], ci.auc(roc_c50)[3], 
               ci.auc(roc_rf)[3], ci.auc(roc_obj)[3])
) %>%
  arrange(desc(AUC)) %>%
  mutate(
    AUC_CI = sprintf("%.3f (%.3f-%.3f)", AUC, CI_lower, CI_upper)
  )

cat("\nModel Performance Comparison:\n")
print(model_comparison %>% select(Model, AUC_CI), row.names = FALSE)

# Plot ROC curves
png("../figures/figure15_ml_roc_comparison.png",
    width = 10, height = 10, units = "in", res = 300)

plot(roc_glm, col = "blue", lwd = 2.5, legacy.axes = TRUE,
     main = "ROC Curves: Model Comparison",
     cex.main = 1.5, cex.lab = 1.3)
plot(roc_c50, col = "green", lwd = 2.5, add = TRUE)
plot(roc_rf, col = "purple", lwd = 2.5, add = TRUE)
plot(roc_obj, col = "yellow", lwd = 2.5, add = TRUE)

abline(a = 0, b = 1, lty = 2, col = "gray50", lwd = 1.5)

legend("bottomright",
       legend = c(
         sprintf("Logistic Regression (AUC = %.3f)", auc(roc_glm)),
         sprintf("C5.0 Decision Tree (AUC = %.3f)", auc(roc_c50)),
         sprintf("Random Forest (AUC = %.3f)", auc(roc_rf)),
          sprintf("LightGBoost (AUC = %.3f)", auc(roc_obj))
       ),
       col = c("blue", "green", "purple", "yellow"),
       lwd = 2.5,
       cex = 1.1,
       bty = "n")

grid(col = "gray", lty = "dotted")

# Save comparison table
write_csv(model_comparison, "../results/ml_model_comparison.csv")
```


### SHAP Values Analysis (Model Explainability)


```{r}
# Prepare data for SHAP
test_data_matrix <- test_data %>%
  select(age_icu_exit, gender_num, log_lactate, 
         log_ck, log_bilirubin, mabp_value_num) %>%
  as.matrix()

train_data_matrix <- train_data %>%
  select(age_icu_exit, gender_num, log_lactate, 
         log_ck, log_bilirubin, mabp_value_num) %>%
  as.matrix()

# Calculate SHAP values for LightGBoost
cat("\nCalculating SHAP values for LightGBoost model...\n")
shap_lgb <- shapviz(
  lgb_model,
  X_pred = test_data_matrix
)

```


```{r}
# Summary plot
png("../figures/figure16_shap_summary.png",
    width = 10, height = 8, units = "in", res = 300)

sv_importance(shap_lgb, kind = "beeswarm")

# Bar plot of mean absolute SHAP values
png("../figures/figure17_shap_importance_bar.png",
    width = 10, height = 7, units = "in", res = 300)

sv_importance(shap_lgb, kind = "bar")

# Dependence plots for top features
png("../figures/figure18_shap_dependence_lactate.png",
    width = 10, height = 7, units = "in", res = 300)

sv_dependence(shap_lgb, v = "log_lactate", color_var = "auto")

# Get SHAP values as dataframe
shap_values_df <- as.data.frame(shap_lgb$S)
colnames(shap_values_df) <- c("age", "gender", "lactate", "ck", "bilirubin", "mabp")

# Calculate mean absolute SHAP values
shap_importance <- colMeans(abs(shap_values_df)) %>%
  sort(decreasing = TRUE)

cat("\nMean Absolute SHAP Values (Feature Importance):\n")
print(round(shap_importance, 4))

# SHAP interpretation
cat("\n\nSHAP INTERPRETATION:\n")
cat(rep("=", 70), "\n")
cat("SHAP values measure each feature's contribution to individual predictions.\n")
cat("Higher |SHAP| = greater impact on mortality prediction.\n\n")

cat("Key insights:\n")
cat(sprintf("1. %s has the highest average impact (%.4f)\n",
            names(shap_importance)[1], shap_importance[1]))
cat(sprintf("2. %s is second most important (%.4f)\n",
            names(shap_importance)[2], shap_importance[2]))
cat(sprintf("3. %s has the least impact (%.4f)\n",
            names(shap_importance)[length(shap_importance)], 
            shap_importance[length(shap_importance)]))

cat("\nConclusion: ", names(shap_importance)[1], 
    " is the most critical predictor across all patients.\n")

```


# ============================================================================
# 8. FEATURE IMPORTANCE COMPARISON ACROSS MODELS
# ============================================================================


```{r}
# Combine importance from all models
importance_combined <- data.frame(
  Feature = c("age_icu_exit", "gender_num", "log_lactate", 
              "log_ck", "log_bilirubin", "mabp_value_num"),
  C50 = c50_importance[c("age_icu_exit", "gender_num", "log_lactate",
                         "log_ck", "log_bilirubin", "mabp_value_num")],
  RandomForest = rf_importance[, "MeanDecreaseGini"],
  LGBoost = importance_lgb$Gain[match(
    c("age_icu_exit", "gender_num", "log_lactate", 
      "log_ck", "log_bilirubin", "mabp_value_num"),
    importance_lgb$Feature
  )],
  SHAP = shap_importance[c("age", "gender", "lactate", "ck", "bilirubin", "mabp")]
)

# Normalize to 0-100
importance_normalized <- importance_combined %>%
  mutate(
    C50 = C50 / max(C50, na.rm = TRUE) * 100,
    RandomForest = RandomForest / max(RandomForest) * 100,
    LGBoost = LGBoost / max(LGBoost) * 100,
    SHAP = SHAP / max(SHAP) * 100
  )

cat("\nNormalized Feature Importance (0-100):\n")
print(round(importance_normalized, 1))

# Save importance comparison
write_csv(importance_normalized, "results/tables/feature_importance_comparison.csv")

# Plot comparison
png("../figures/figure19_importance_comparison.png",
    width = 12, height = 8, units = "in", res = 300)

importance_long <- importance_normalized %>%
  pivot_longer(cols = c(C50, RandomForest, LGBoost, SHAP),
               names_to = "Model",
               values_to = "Importance") %>%
  mutate(
    Feature = case_when(
      Feature == "age_icu_exit" ~ "Age",
      Feature == "gender_num" ~ "Gender",
      Feature == "log_lactate" ~ "Lactate",
      Feature == "log_ck" ~ "CK",
      Feature == "log_bilirubin" ~ "Bilirubin",
      Feature == "mabp_value_num" ~ "MAP"
    )
  )
```




```{r}
library(ggplot2)
ggplot(importance_long, aes(x = reorder(Feature, Importance), 
                            y = Importance, fill = Model)) +
  geom_col(position = "dodge", alpha = 0.8, color = "black") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Feature Importance Comparison Across Models",
    subtitle = "Normalized to 100 for comparison",
    x = "Feature",
    y = "Relative Importance (%)",
    fill = "Model"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

dev.off()

cat("\nFigure saved: results/figures/figure19_importance_comparison.png\n")
```

# ============================================================================
# SAVE ALL RESULTS
# ============================================================================

```{r}
ml_results <- list(
  model_glm = model_glm,
  model_c50 = model_c50,
  model_c50_rules = model_c50_rules,
  model_rf = model_rf,
  model_xgb = model_xgb,
  model_comparison = model_comparison,
  importance_comparison = importance_normalized,
  shap_values = shap_values_df,
  test_predictions = data.frame(
    actual = test_data$icu_dead_f,
    pred_glm = pred_glm,
    pred_c50 = pred_c50,
    pred_rf = pred_rf,
    pred_xgb = pred_xgb
  )
)

save(ml_results, file = "results/ml_complete_results.RData")

cat("\n", rep("=", 70), "\n")
cat("MACHINE LEARNING ANALYSIS COMPLETE\n")
cat(rep("=", 70), "\n")
cat("\nBest performing model: ", 
    model_comparison$Model[which.max(model_comparison$AUC)], "\n")
cat(sprintf("AUC: %.3f\n", max(model_comparison$AUC)))
cat("\nMost important feature (SHAP): ", names(shap_importance)[1], "\n")
cat("\nAll results saved to: results/ml_complete_results.RData\n")
```


***






------------------------------------------------------------------------

## Subgroup analyses, sensitivity analyses

## Create all figures and tables


```{r}
str(eda_df)
```

```{r}
if (!require('cluster')) install.packages('cluster')
library(cluster)

if (!require('dbscan')) install.packages('dbscan')
library(dbscan)
```



```{r}
#cols_ml <- c()

imputed_df <- na.omit(initial_df)
summary(imputed_df)
```


```{r}
names(imputed_df)
```

```{r}
ml_columns <- c("subject_id", "age_icu_entry","gender","icu_dead","lactate_time",
                "lactate_value_num","lactate_flag", "ck_value_num","ck_flag",
                "bilirubin_value_num","bilirubin_flag","mabp_value_num","mabp_units"
)
stats_df <- imputed_df %>%
  select(ml_columns)
``` 


```{r}
head(stats_df)
```

```{r}
# Proportion table by gender 0=F 1=M
prop.table(table(ifelse(stats_df$gender == 'F', 0, 1)))*100
```


```{r}
# Proportion table by lactate_flag 0=normal, 1=abnormal
prop.table(table(ifelse(stats_df$lactate_flag == 'normal', 0, 1)))*100
```


```{r}
# Proportion table by bilirubin_flag 0=normal, 1=abnormal
prop.table(table(ifelse(stats_df$bilirubin_flag == 'normal', 0, 1)))*100
```


```{r}
# Proportion table by ck_flag 0=normal, 1=abnormal
prop.table(table(ifelse(stats_df$ck_flag == 'normal', 0, 1)))*100
```


```{r}
# Proportion table by icu_dead 0=alive, 1=dead
prop.table(table(stats_df$icu_dead))*100
```



```{r}
stats_df$lactate_scores <- case_when(
  stats_df$lactate_value_num < 2 ~ 0,
  stats_df$lactate_value_num >=2 & stats_df$lactate_value_num < 4 ~ 1,
  stats_df$lactate_value_num >= 4 ~ 2
)
```

```{r}
set.seed(17)
```


```{r}
# Model 1: Lactate only (unadjusted)
model1 <- glm(
  icu_dead ~ lactate_value_num,
  data = stats_df,
  family = binomial()
)
```



```{r}
# Model 2: Lactate + demographics

stats_df$gender_num <- ifelse(stats_df$gender == 'F', 0, 1)

model2 <- glm(
  icu_dead ~ lactate_value_num + age_icu_entry + gender_num,
  data = stats_df,
  family = binomial()
)
```



```{r}
# Model 3: Lactate + demographics + clinical factors
model3 <- glm(
  icu_dead ~ lactate_value_num + age_icu_entry + gender_num + 
    ck_value_num + bilirubin_value_num + mabp_value_num,
  data = stats_df,
  family = binomial()
)
```


```{r}
# Compare models
cat("\n--- Model 1: Unadjusted ---\n")
print(tidy(model1, exponentiate = TRUE, conf.int = TRUE))
cat(sprintf("AIC: %.1f\n", AIC(model1)))

cat("\n--- Model 2: Adjusted for demographics ---\n")
print(tidy(model2, exponentiate = TRUE, conf.int = TRUE))
cat(sprintf("AIC: %.1f\n", AIC(model2)))

cat("\n--- Model 3: Adjusted for demographics + clinical factors ---\n")
print(tidy(model3, exponentiate = TRUE, conf.int = TRUE))
cat(sprintf("AIC: %.1f\n", AIC(model3)))
```

```{r}
# Likelihood ratio tests
lrt_2vs1 <- anova(model1, model2, test = "Chisq")
lrt_3vs2 <- anova(model2, model3, test = "Chisq")
```



```{r eval=FALSE, include=FALSE}
cat("\nLikelihood Ratio Tests:\n")
cat(sprintf("Model 2 vs Model 1: p = %.4f\n", lrt_2vs1$`Pr(>Chi)`[2]))
cat(sprintf("Model 3 vs Model 2: p = %.4f\n", lrt_3vs2$`Pr(>Chi)`[2]))
```


```{r eval=FALSE, include=FALSE}
# Save model results
model_comparison <- data.frame(
  Model = c("Model 1: Unadjusted",
            "Model 2: + Demographics",
            "Model 3: + Clinical factors"),
  AIC = c(AIC(model1), AIC(model2), AIC(model3)),
  BIC = c(BIC(model1), BIC(model2), BIC(model3))
)
```


```{r eval=FALSE, include=FALSE}
model_comparison
```


```{r eval=FALSE, include=FALSE}
cat("\n=== CATEGORICAL LACTATE ANALYSIS ===\n")

# Using lactate categories (reference: Normal)
model_categorical <- glm(
  icu_dead ~ lactate_scores + age_icu_entry + gender_num + 
    ck_value_num + bilirubin_value_num + mabp_value_num,
  data = stats_df,
  family = binomial()
)

cat("\nLogistic Regression: Lactate Categories (Adjusted)\n")
results_cat <- tidy(model_categorical, exponentiate = TRUE, conf.int = TRUE)
print(results_cat)

```

```{r eval=FALSE, include=FALSE}
cat("\n=== ROC CURVE ANALYSIS ===\n")

# Prepare data for ROC analysis (complete cases only)
roc_data <- stats_df %>%
  mutate(
    pred_model1 = predict(model1, newdata = ., type = "response"),
    pred_model3 = predict(model3, newdata = ., type = "response"),
  )

# Calculate ROC curves
roc_model1 <- roc(roc_data$icu_dead, roc_data$pred_model1)
roc_model3 <- roc(roc_data$icu_dead, roc_data$pred_model3)
```

```{r eval=FALSE, include=FALSE}
# AUC with confidence intervals
cat("\nAUC (95% CI):\n")
cat(sprintf("Model 1 (Lactate only): %.3f (%.3f-%.3f)\n",
            auc(roc_model1), ci.auc(roc_model1)[1], ci.auc(roc_model1)[3]))
cat(sprintf("Model 3 (Clinical model): %.3f (%.3f-%.3f)\n",
            auc(roc_model3), ci.auc(roc_model3)[1], ci.auc(roc_model3)[3]))

# Compare ROC curves
roc_comparison <- roc.test(roc_model1, roc_model3)
cat(sprintf("\nROC comparison (Model 3 vs Model 1): p = %.4f\n", 
            roc_comparison$p.value))

# Create Figure 1&3: ROC Curves
png("../figures/figure4_roc_curves.png", 
    width = 10, height = 10, units = "in", res = 300)

plot(roc_model1, col = "blue", lwd = 2.5, legacy.axes = TRUE,
     main = "ROC Curves for Mortality Prediction",
     cex.main = 1.5, cex.lab = 1.3, cex.axis = 1.2)
plot(roc_model3, col = "red", lwd = 2.5, add = TRUE)

legend("bottomright",
       legend = c(
         sprintf("Lactate only (AUC = %.3f)", auc(roc_model1)),
         sprintf("+ Clinical factors (AUC = %.3f)", auc(roc_model3))
       ),
       col = c("blue", "red", "darkgreen"),
       lwd = 2.5,
       cex = 1.2,
       bty = "n")

grid(col = "gray", lty = "dotted")

dev.off()
```



```{r calibration analysis, eval=FALSE, include=FALSE}
cat("\n=== CALIBRATION ANALYSIS ===\n")

# Use rms package for proper calibration
dd <- datadist(roc_data)
options(datadist = "dd")

# Fit model using lrm
model_rms <- lrm(
  icu_dead ~ lactate_value_num + age_icu_entry + gender_num + 
    ck_value_num + bilirubin_value_num + mabp_value_num,
  data = stats_df,
  family = binomial(),
  data = roc_data,
  x = TRUE,
  y = TRUE
)
```



```{r boostrap_validatiom, eval=FALSE, include=FALSE}

# Bootstrap validation
set.seed(17)
val <- validate(model_rms, B = 200)

cat("\nBootstrap Validation (200 repetitions):\n")
print(val)

# Calibration
cal <- calibrate(model_rms, B = 200)

# Plot calibration
png("../figures/figure5_calibration.png",
    width = 8, height = 8, units = "in", res = 300)

plot(cal,
     xlab = "Predicted Probability of Mortality",
     ylab = "Observed Probability of Mortality",
     main = "Calibration Plot (200 Bootstrap Samples)",
     cex.main = 1.4,
     cex.lab = 1.2,
     cex.axis = 1.1)

dev.off()

```



------------------------------------------------------------------------

\newpage

# Next steps

------------------------------------------------------------------------

\newpage

# Conclusions

## Final thoughts

------------------------------------------------------------------------

\newpage

# References

------------------------------------------------------------------------
